Cosmic Microwave Background Power Spectra from Few Bit Timestreams

### 2018-09-17 digitisation.tex
Included 1/f noise, editing, updated plots

\usepackage[table,usenames,dvipsnames]{xcolor}
%\usepackage{amsmath}
\usepackage{subfigure}
\usepackage[backref,breaklinks,colorlinks,citecolor=blue]{hyperref}
\usepackage{natbib}
%\usepackage{natbib}
\bibliographystyle{fapj}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{soul}

@@

\newcommand{\cpmftens}{\ensuremath{C_{\ell}^{\rm PMF,\,tens}}}
\newcommand{\cpmfvec}{\ensuremath{C_{\ell}^{\rm PMF,\,vec}}}
\newcommand{\apmf}{\ensuremath{A_{\rm PMF}}}
\newcommand{\bpmf}{\ensuremath{B_{\rm 1\,Mpc}}}
\newcommand{\alens}{\ensuremath{A_{\rm lens}}}
\newcommand{\lcdm}{\ensuremath{\Lambda}CDM}
\newcommand{\nrun}{\ensuremath{n_{\rm run}}}
\newcommand{\neff}{\ensuremath{N_{\rm eff}}}
\newcommand{\ho}{H\ensuremath{_0}}
\newcommand{\mnu}{\ensuremath{\sum m_\nu}}
\newcommand{\ukarcmin}{\ensuremath{\mu}{\rm K-arcmin}}
\newcommand{\lknee}{\ensuremath{\ell_{\rm knee}}}
\newcommand{\fermilat}{\textit{Fermi}-LAT}

@@

Observations of the Cosmic Microwave Background (CMB) are of significant value to modern cosmology and particle physics. Future CMB experiments face issues in hardware design, mission planning and analysis that arise from the vast size of time-ordered-data (TOD) being recorded. These challenges are particularly significant for Antarctica- and space-based experiments which depend on satellite links to transmit their data. We explore the viability of reducing the TOD to few-bit numbers to address these issues. Unlike lossless compression, this technique introduces additional noise into the data. We present 1, 2 and 3 bit digitisation schemes and determine the degradation this compression causes in temperature and polarisation power spectra. We find that 3 bit digitisation has a percent-level contribution to the map noise level. We argue that such digitisation is a promising strategy for upcoming experiments.

@@

Observations of the Cosmic Microwave Background (CMB) have played a key role in physics since its discovery \citep{penzias1965}. Current and future CMB experiments will continue to deliver new insights by studying the temperature and polarisation information contained in the CMB. These will put tight constraints on cosmological models. Upcoming science goals include the discovery of inflationary gravitational waves, construction of precise lensing maps and thorough studies of the Sunyaev-Zeldovich (SZ) effects. Moreover, CMB experiments will measure the relativistic number of species and the neutrino mass sum \citep{s4sciencebook}. %L: reference to science goals by satellite missions as well

High fidelity CMB science demands an excellent standard of data analysis. The CMB community has developed a variety of compression techniques and computational approaches to manage the increasing influx of data, while maximising the science output \citep{tristam2007}. These include the compression of time-ordered data (TOD) into maps \citep{tegmark1997}, bandpower estimation \citep{tegmark1998} and the pseudo $C_l$ method \citep{brown2005}.

A growing hurdle for experiments at remote locations are the transmission limitations of satellite links. Space-based experiments have employed a combination of lossless and lossy compression techniques, including reduced bits in the TOD \citep{gaztanaga1998, maris2003}. Antarctica-based experiments that transmit a portion of their data via a satellite link have downsampled their data in the past to meet their telemetry requirements. They have yet to exploit few bit digitisation of the TOD. As we approach the next generation of ground-based experiments, Stage-4, and the launch of a new generation of space-based missions (liteBIRD, PIXIE, COrE+), we must treat the transmission bottleneck carefully. Without a thorough review of compression techniques we are sure to lose information. %L:ref to non-planck satellites for reduced bit TOD? %C: Is there a reference detailing downsampling of SPT data?

In this work we present the method of extreme digitisation, which compresses a rich digital input signal into a few bits. We apply extreme digitisation to the TOD and detail its effect on temperature and polarisation power spectra. We find that an optimal 3-bit digitisation scheme adds as little as $< 2\%$ to the map noise level. The digtisation schemes described here are laid out for ground-based experiments. Space-based missions must study carefully the effect of digitisation in their specific compression algorithms.% of t the results we present here should also be considered by future space-based missions, which inevitably must incorporate lossy compression.

This work is structured as follows. In \S\ref{sec:dig} we detail the challenges originating from handling large TOD, introduce the process of extreme digitisation and lay out the framework used to test its performance. Subsequently, we describe the power spectrum estimation employed and interpret the obtained results in section \S\ref{sec:results}. We summarise our findings in \S\ref{sec:conclusions}.

@@

\subsection{Problem}
\label{subsec:problem}

@@

The science goals of upcoming CMB experiments naturally lead to a large influx of data To achieve the targeted sensitivity longer observations with more detectors are needed. In fact the number of detectors of ground-based experiments has been following a Moore's law like trend, doubling approximately every 2 years. This directly translates into an exponential growth in data volume \citep{s4sciencebook}. %L: check out the particle paper where the moores law figure comes from

Excellent observation conditions for CMB observations can not only be found in space, but also at various locations on earth \citep{li2017, kovac2007} among which is the South Pole. Space- and Antarctica-based experiments depend on satellite transmission. An experiment at the South Pole will be part of the next generation of ground-based instruments, which aims to collect 2 million detector years in total. Such an experiment will face a data influx of $\sim \mathrm{O}(10)\mathrm{Tb/d}$. However, the current transmission allocation for SPT3G is at $150\mathrm{Gb/d}$, which will likely only see a moderate increase in coming years. The transmission bottleneck is currently overcome by recovering the full data on hard drives with some latency and by transmitting a downsampled version of the data. The downsampling process loses high frequency information. Going into Stage-4 we anticipate that compression rates for transmission must increase by an order of magnitude. Continued use of downsampling will narrow the information window decisively - prohibiting high multipole moment science to be carried out on the transmitted data set. This also means that any potential faults or errors in the experiment that only become visible in said range will go unnoticed for longer. %L: reference for observing conditions in chile or a general review of sites %C: Is there a paper detailing trasmission limits for SPT and that the data is currently downsampled?

@@

Beyond transmission challenges, mission planning is becoming exceedingly difficult. A full simulation of TOD over the entire parameter space of detection scenarios for numerous set-ups is the desired way to decide on Stage-4 configurations. Space-based missions must aim to carry out a similar analysis to optimise their science output. Given the shear size of TOD expected, this is not possible \citep{s4sciencebook}. We must rely on different planning strategies or aim to reduce the size of the TOD in order to maximise the productivity of planning and development stages and guarantee scientific excellence.

@@

Operations on the TOD, such as noise-removal or map-making are a vital part of CMB data analysis. While we have not experienced the limitations of the accessible computational assets, the exponential growth of CMB data makes its analysis increasingly expensive. The community depends on the facilities provided by the National Energy Research Scientific Computing Center (NERSC).

@@

Extreme Digitisation would tackle the challenges mentioned above by reducing the size of the TOD by an order of magnitude. Together with already established lossless compression techniques (e.g. FLAC, run-length coding, Huffman coding, etc.) this will directly tackle transmission hurdles. Beyond that extreme digitisation has the potential of easing issues in planning, analysis and hardware requirements.

The Planck mission has demonstrated the advantages of using few bit compression of the TOD \citep{maris2003}. Other science areas have also shown the power of extreme digtisation. \citep{jenet1998} explored the application of such compression to radio pulsar timing measurements with success. Recently \citep{clearwater2018} have investigated the advantages of using 1 and 2 bit data when searching for continuous gravitational waves using the Laser Interferometer Gravitational-Wave Observatory (LIGO).

@@

Digitisation is a lossy compression technique. However, the induced noise depends on the number of bits used, the digitisation thresholds, and the output levels chosen. To minimise the noise induced through this process one must know the nature of the input signal. There is usually little concern around finding an optimal set of digitisation parameters for a given input signal. It is common to have access to large numbers of bits to store information, where changes in the digitisation scheme become insignificant to the distortion induced. It is in our interest however to consider extreme, i.e. few-bit, digitisation, where a digitisation scheme must be carefully constructed. We review the key ideas regarding construction of digitisation schemes as laid out by \cite{max1960} below.

@@

where $s$ is the input and $\hat{s}$ the output signal. For an input signal that has at least some stochastic element to it we introduce the input amplitude probability density $p(x)$. This allows us to rewrite the above as

@@

To progress further we have to make an assumption about the distribution of input signals, $p(x)$. For our purposes we assume that CMB observations operate at low signal to noise. Furthermore we assume that the noise profile is Gaussian white noise with standard deviation $\sigma$, i.e. $p(x) = (1/\sqrt{2\pi\sigma^2}) e^{-x^2/2\sigma^2}$. Given this assumption we can solve the problem using a numerical iterative procedure. One begins by picking $y_1$ and calculating the remaining $x_i$'s and $y_i$'s using equation \ref{eq:digitequalspacecondition}. Afterwards one checks whether this choice of values satisfy the conditions given by equation \ref{eq:digitareacondition}. If that is the case, the $x_i$'s and $y_i$'s were chosen appropriately.

@@

\begin{equation} \hat{s}_1(t) = \left\{ \begin{array}{lr}
1, & \text{for } s(t) > 0\\
-1, & \text{for } s(t) \leq 0
\end{array} \right. \end{equation}

@@

\begin{equation} \hat{s}_2(t) = \left\{ \begin{array}{rl}
1.51 \sigma, & \text{for } s(t) \geq 0.9816 \sigma\\
0.4528 \sigma, & \text{for } 0 \leq s(t) < 0.9816 \sigma\\
-0.4528 \sigma, & \text{for } 0.9816 \sigma \leq s(t) < 0\\
-1.51 \sigma, & \text{for } 0.9816 \sigma < s(t)\\
\end{array} \right. . \end{equation}
%C: should i round to equal significant figures here? Unfortunately these are the exact values that i used...

@@

\begin{equation} \hat{s}_3(t) = \left\{ \begin{array}{rl}
2.152 \sigma, & \text{for } s(t) \geq 1.748 \sigma\\
1.344 \sigma, & \text{for } 1.05 \sigma \leq s(t) < 1.748 \sigma\\
0.756 \sigma, & \text{for } 0.501 \sigma \leq s(t) < 1.05 \sigma\\
0.245 \sigma, & \text{for } 0 \leq s(t) < 0.501 \sigma\\
-0.245 \sigma, & \text{for } 0.501 \sigma \leq s(t) < 0\\
-0.756 \sigma, & \text{for } 1.05 \sigma \leq s(t) < 0.501 \sigma\\
-1.344 \sigma, & \text{for } 1.748 \sigma \leq s(t) < 1.05 \sigma\\
-2.152 \sigma, & \text{for } 1.748 \sigma < s(t)\\
\end{array} \right. . \end{equation}

The fact that the single bit digitisation scheme makes no references to the standard deviation of the signal is irrelevant. This affects the normalisation of the signal, which we can account for later. What is important is the spacing between digitisation thresholds and relative distance between the output levels chosen.

Completely different digitisation schemes can be thought of, which would for example reference the most drastic outlier in the data set, or seek to place equal numbers of points into each digitisation level. However, given the assumptions made the schemes derived above are optimal. Additionally, they are simple enough to be easily implemented computationally.

@@

To investigate the performance of the derived digitisation schemes we simulate many scans over CMB template maps at the timestream level. Each scan is performed by a single detector. We obtain control maps that use 64bit TOD and maps that have undergone 1, 2 and 3 bit digitisation at the timestream level. We calculate the temperature and polarisation power spectra of each map and determine the additional noise induced through the extreme digitisation process.
%C: Do I need to mention that we assume to have an infinitely fast spinning plate? %L: Check whether Python uses 32 or 64 bit under on NERSC for floats

To create the template maps we use the Hierarchical Equal Area isoLatitude Pixelization of a sphere \citep[HEALPix;][]{healpix}\footnote{Available at \url{http://healpix.sourceforge.net/}} framework. We generate a realisation of I, Q and U maps with $\mathrm{NSIDE} = 4096$ based on the results of the Planck collaboration with key cosmological parameters of $\Omega_b \approx 0.049 $, $\Omega_c \approx 0.265$, $\Omega_m \approx 0.316$ and $h \approx 0.67$ \citep{planck2016}.

@@

We simulate observing a $\sim 600 \mathrm{deg^2}$ patch of the sky. To do so we perform a number of constant elevation scans (CES), equally spaced in declination (DEC). We repeat the observation strategy $100$ times with a slight offset in right ascension (RA) and DEC each time, such that each pixel is sampled approximately uniformly. The speed at which we sweep across the survey area is adjusted to produce a desired number of hits per pixel in the output maps.

While performing each CES the pixels being targeted are determined. The corresponding values from the template maps are then accessed and added to realisations of the detector noise of appropriate length to generate TOD streams. We assume the detector noise to be Gaussian white noise. For temperature observations we assume the detector noise level to be $500 \mathrm{\mu K \sqrt{s}}$ and for polarisation observations $\sqrt{2} \hspace{0.7ex} 500 \mathrm{\mu K \sqrt{s}}$, i.e. the corresponding photon noise limit.

At this point we apply the digitisation schemes to the TOD. We compress the timestream into maps by averaging all hits falling into the same pixel. We produce 12 CMB maps in total: three control maps (I, Q, U) and nine maps with three each obtained from each digitisation scheme.

@@

This process is carried out 6 times. We produce maps where pixels in the survey patch are hit approximately uniformly $800$, $8,000$, $80,000$, $1,024,000$, $10,240,000$ and $102,400,000$ times. We assume that the detector is read out $200$ times per second. %The calculation outlined above has considerable computational requirements if we want to reach up to $\sim 10^8$ hits per pixel. This calculation was made possible through use of the NERSC facilities.

%\begin{table}[tbh]
%\begin{center}
%\caption{\label{tab:inputcosparams} Input Cosmological Parameters}
%\small
%\begin{tabular}{l | c}
%Parameter & Planck 2015\\
%\hline\\begin{equation}-1.5ex]
%$100\theta_{MC}$ & $ 1.04086 \pm 0.00048$\\
%$\Omega_b h^2$ & $  0.02222 \pm 0.00023$\\
%$\Omega_c h^2$ & $ 0.1199 \pm 0.0022$\\
%$H_0$ & $67.26 \pm 0.98$\\
%$n_s$ & $0.9652 \pm 0.0062$\\
%$\Omega_m$ & $0.316 \pm 0.014$\\
%$\sigma_8$ & $0.830 \pm 0.015 $\\
%$\tau$ & $0.078 \pm 0.019$\\
%$10^9 A_s e^{-2\tau}$ & $1.881 \pm 0.014 $\\
%\end{tabular}
%\tablecomments{ 
%Cosmological parameters used to create the template maps. Taken from Planck 2015: Cosmological Parameters.
%} \normalsize
%\end{center}
%\end{table}

%\begin{table}[tbh] % transpose and include hits per pixel!
%\begin{center}
%\caption{\label{tab:modelparams} Assumed Survey Parameters}
%\small
%\begin{tabular}{l | c}
%Parameter & Value\\
%\hline \\begin{equation}-1.5ex]
%$\mathrm{NSIDE}$ &$4096$\\
%$f_{\mathrm{readout}} [\mathrm{Hz}]$ &$200$\\
%$f_{{\mathrm{sky}}}$ & $\sim 0.014$\\
%$\sigma^{\mathrm{T}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ &$500$\\
%$\sigma^{\mathrm{Pol}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ & $\sqrt{2}\times500$\\
%\multirow{2}{*}{$\mathrm{hits per pixel}$} & $\left(1, 10, 100 \right) \times 800,$\\
%& $\left( 1, 10, 100 \right) \times 1024000$ \\
%\end{tabular}
%\tablecomments{ 
%Parameters used in the simulated observation. The RA speed is adjusted to match the desired hits per pixel.
%} \normalsize
%\end{center}
%\end{table}

@@

We use the Spatially Inhomogeneous Correlation Estimator for Temperature and Polarisation \citep[PolSpice;][]{polspice} to compute the $\mathrm{TT}, \mathrm{EE}$ and $\mathrm{BB}$ power spectra of the reconstructed maps I, Q, U maps. To minimise cut sky effects we apodise the observed path using a cosine mask with $\sigma_{\mathrm{APOD}} = \sqrt{600}/2 \hspace{0.7ex} \mathrm{deg}$. %C: Should I also mention the multiplication with a cosine mask that I supply in addition to setting this option on polspice? This will likely be reworked when using Healpix to get the power spectra though.

%%%%%%%%%%%
% INCLUDE SECTION OF ALM CUTS (maybe a plot of the cuts too)
%%%%%%%%%%%

%\begin{table}[tbh]
%\begin{center}
%\caption{\label{tab:polspiceparams} PolSpice Parameters}
%\small
%\begin{tabular}{c c c c}
%weights & $\sigma_{\mathrm{APOD}} [\mathrm{deg}]$ & apodistation type & polarisation \\
%\hline
%cosine mask & $\sqrt{600}/2$ & cosine & Yes \\
%
%\end{tabular}
%\tablecomments{ 
%Parameters used when calling PolSpice to calculate the power spectra. Remaining parameters have been left at their default value.
%} \normalsize
%\end{center}
%\end{table}

@@

Normalisation of the obtained power spectra is necessary. The digitisation schemes introduced in section \S\ref{sec:dig} are designed to minimise the distortion of the TOD, not to preserve power at the map level. Normalising obtained spectra against previous experiments is a common calibration technique in many CMB experiments (e.g. SPT). %C: Do you know a good reference for this?

For the normalisation procedure we focus on the $\sim 10^8$ hits per pixel maps. We normalise the $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra from few bit TOD against their control counterparts. We consider the normalisation window cut out by the conditions

\begin{equation} l > 35 \> , \> \frac{C_l^{\mathrm{S}}}{C_l^{\mathrm{N}}} \geq 10. \end{equation}

where $C_l^{\mathrm{S}}$ is the power spectrum of the input template maps and $C_l^{\mathrm{N}}$ the detector noise level. The lower limit is placed in accordance to the patch-size surveyed while the upper limit ensures appropriate signal to noise in the normalisation window.

The normalisation constants for each digitisation scheme and channel obtained through this way are then applied to all lower hits per pixel simulations. A sample of the obtained power spectra is shown in figure X.

@@

%\begin{figure}[htb]\centering
%\includegraphics[width=0.5\textwidth,clip]{Plots/ttzoom.png}
%\includegraphics[width=0.5\textwidth,clip]{Plots/eezoom.png}
%\includegraphics[width=0.5\textwidth,clip]{Plots/bbzoom.png}
%  \caption[Current ]{
%   Reconstructed $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra. These originate from an observation with $102400000 \mathrm{hits per pixel}$. We observe that digitisation appears to add some constant to the noise level. As one would expect $3 \mathrm{Bit}$ digitisation performs the best, followed by $2 \mathrm{Bit}$- and finally $1 \mathrm{Bit}$ digitisation.
%\label{fig:powspeczoom}
%}
%\end{figure}

\subsection{Additional Noise}
\label{subsec:additionalnoise}

%C: This is where I'm planning on
%%%%%%%%%%%
% INSERT A COUPLE PARAGRAPHS ABOUT THE PS OF DIFFERENCE MAPS
% INCLUDE PLOTS OF PS OF DIFFERENCE MAPS
% want to show that beyond some calibration error, we do not observe any l dependence in the added noise
%%%%%%%%%%%

To quantify the distortion caused by extreme digitisation we infer the map noise level from the relevant power spectra. These are put into the context of the map noise level deduced from the control power spectra. We formulate

@@

In this work we have motivated the investigation of extreme digitisation as a technique in combating arising challenges in CMB data analysis. The reduction of TOD by an order of magnitude directly addresses issues in transmission faced by remote location observations. Benefits in mission planning, data analysis and hardware requirements are possible.

>>>>

\usepackage[table,usenames,dvipsnames]{xcolor}
\usepackage{amsmath}
\usepackage{subfigure}
\usepackage[backref,breaklinks,colorlinks,citecolor=blue]{hyperref}
\usepackage{natbib}
%\usepackage{natbib}
\bibliographystyle{fapj}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{soul}

@@

\newcommand{\cpmftens}{\ensuremath{C_{\ell}^{\rm PMF,\,tens}}}
\newcommand{\cpmfvec}{\ensuremath{C_{\ell}^{\rm PMF,\,vec}}}
\newcommand{\apmf}{\ensuremath{A_{\rm PMF}}}
\newcommand{\bpmf}{\ensuremath{B_{\rm 1\,Mpc}}}
\newcommand{\alens}{\ensuremath{A_{\rm lens}}}
\newcommand{\lcdm}{\ensuremath{\Lambda}CDM}
\newcommand{\nrun}{\ensuremath{n_{\rm run}}}
\newcommand{\neff}{\ensuremath{N_{\rm eff}}}
\newcommand{\ho}{H\ensuremath{_0}}
\newcommand{\mnu}{\ensuremath{\sum m_\nu}}
\newcommand{\ukarcmin}{\ensuremath{\mu}{\rm K-arcmin}}
\newcommand{\lknee}{\ensuremath{\ell_{\rm knee}}}
\newcommand{\lcut}{\ensuremath{\ell_{\rm cut}}}
\newcommand{\fermilat}{\textit{Fermi}-LAT}

@@

Observations of the Cosmic Microwave Background (CMB) are of significant value to modern cosmology and particle physics. The large size of time-ordered-data (TOD) being recorded causes new challenges in hardware design, mission planning and analysis for future CMB experiments. These issues are particularly significant for Antarctica- and space-based experiments which depend on satellite links to transmit their data. We explore the viability of reducing the TOD to few-bit numbers to address these issues. Unlike lossless compression, this technique introduces additional noise into the data. We present a set of 1, 2 and 3 bit digitisation schemes and determine the degradation the compression causes in temperature and polarisation power spectra. The added noise has no dependence on the multipole moment $\ell$ for white and $1/f$ detector noise. We note that 3 bit digitisation has a percent-level contribution to the map noise level in all cases considered. Such digitisation is a promising strategy for upcoming experiments.

@@

Observations of the Cosmic Microwave Background (CMB) have played a key role in physics since its discovery \citep{penzias1965}. Current and future study of the information it contains will put tight constraints on cosmological models. Upcoming science goals include the discovery of inflationary gravitational waves, construction of precise lensing maps and thorough studies of the Sunyaev-Zeldovich (SZ) effects. Additionally, the CMB probes structure formation and reionisation. It is a powerful test for dark matter and dark energy models. Lastly, the relativistic number of species and the neutrino mass sum are measured by CMB experiments \citep{s4sciencebook, core2011, litebird2016, pixie2011}.

The CMB community has developed a variety of compression techniques and computational approaches to handle the increasing volume of data the science goals demand \citep{tristam2007}. These include the compression of time-ordered data (TOD) into maps \citep{tegmark1997}, bandpower estimation \citep{tegmark1998} and the pseudo $C_l$ method \citep{brown2005}.

A potential hurdle for experiments at remote locations are the transmission limitations of satellite links. Space-based experiments have employed a combination of lossless and lossy compression techniques, including reduced bits in the TOD \citep{gaztanaga1998, maris2003}. Antarctica-based experiments that transmit a portion of their data via a satellite link have downsampled their data in the past to meet their telemetry requirements and have not yet to used few bit digitisation of the TOD. As we approach the next generation of ground-based experiments, CMB-S4, and the launch of a new generation of space-based missions (LiteBIRD, PIXIE, COrE+), we must consider potential transmission bottlenecks carefully. %C: Is there a reference detailing downsampling of SPT data?

In this work we present the method of extreme digitisation, which reduces a many bit (often 32-or 64-bit) signal to a few bits for ground-based experiments. We apply extreme digitisation to the TOD and detail the resulting effects on temperature and polarisation power spectra. We find that an optimal 3-bit digitisation scheme adds $< 2\%$ to the map noise level.

This work is structured as follows. In \S\ref{sec:dig} we detail the challenges that come with handling large data volumes, introduce the process of extreme digitisation and lay out the framework used to test its performance. We describe the power spectrum estimation and present the results in section \S\ref{sec:results}. We summarise our findings in \S\ref{sec:conclusions}.

@@

\subsection{The Challenges Of Large Data Sets}
\label{subsec:problem}

@@

The science goals of upcoming CMB experiments depend on achieving substantially faster mapping speeds. This has lead to the number of detectors of ground-based experiments following a Moore's law like trend, doubling approximately every 2 years \citep{s4sciencebook, Abazajian2015}. The growing detector count translates into an exponential growth in data volume.

Excellent observation conditions for CMB observations can be found at various locations on earth, including the South Pole \citep{kovac2007}, and in space. Space- and Antarctica-based experiments depend on satellite transmission. An experiment at the South Pole will be part of the next generation of ground-based instruments, which aims to collect 2 million detector years in total. Such an experiment will face a data influx of $\sim \mathcal{O}(10)\mathrm{Tb/d}$. However, the current transmission allocation for SPT3G is at $150\mathrm{Gb/d}$, which will likely see only a moderate increase in coming years. The transmission bottleneck is currently overcome by recovering the full data on hard drives with some latency and transmitting a downsampled version of the data. The downsampling process loses high frequency information. Continued use of downsampling will narrow the information window decisively - prohibiting high multipole moment science to be carried out on the transmitted data set. This also means that any potential faults or errors in the experiment visible in this range are not visible in the transmitted data. %C: Is there a paper detailing trasmission limits for SPT and that the data is currently downsampled?

@@

Beyond transmission challenges, mission planning is becoming exceedingly difficult. Full simulation of TOD over the entire parameter space of detection scenarios is the desired way to decide on Stage-4 configurations. Given the shear size of TOD expected, this is not possible \citep{s4sciencebook}. Space-based missions carry out a similar analysis specific to their instruments to optimise their science output. Ground-based observations have to rely on different planning strategies or aim to reduce the size of the TOD in order to maximise the productivity of instruments.

@@

Operations on the TOD, such as noise-removal or map-making are a vital part of CMB data analysis. While we have not experienced the limitations of the accessible computational assets, the exponential growth of CMB data makes its analysis increasingly expensive.

@@

Extreme Digitisation would tackle the challenges mentioned above by reducing TOD volume by an order of magnitude. Together with already established lossless compression techniques (e.g. FLAC, run-length coding, Huffman coding, etc.) this will directly tackle transmission hurdles. Beyond that extreme digitisation has the potential of easing issues in planning, analysis and hardware requirements.

The Planck mission has demonstrated the advantages of using few bit compression of the CMB data \citep{maris2003}. Other science areas have also shown the power of extreme digtisation. \cite{jenet1998} explored the application of such compression to radio pulsar timing measurements with success. Recently Clearwater et al. (private communication) have investigated the advantages of using 1 and 2 bit data when searching for continuous gravitational waves using the Laser Interferometer Gravitational-Wave Observatory (LIGO).

@@

Digitisation is a lossy compression technique. However, the induced noise depends on the details of the digitisation scheme. To minimise the noise induced through this process one must know the nature of the input signal. There is usually little concern around finding an optimal set of digitisation parameters for an input signal as it is common to have access to large numbers of bits to store information, where changes in the digitisation scheme become insignificant to the distortion induced. It is in our interest however to consider extreme, i.e. few-bit, digitisation. In this case the digitisation scheme must be carefully constructed. We review the key of digitisation as edtablished by \cite{max1960} below.

@@

where $s$ is the input and $\hat{s}$ the output signal. For an input signal that has a stochastic element we introduce the input amplitude probability density $p(x)$. This allows us to rewrite the above as

@@

To progress further we have to make an assumption about the probability distribution of input signals. We suppose CMB observations are low signal to noise and detector noise is Gaussian white. The input is then purely stochastic and  $p(x) = (1/\sqrt{2\pi\sigma^2}) e^{-x^2/2\sigma^2}$, where $\sigma$ is the standard deviation of the detector noise. Under these assumptions the system of equations can be solved numerically. One begins by picking $y_1$ and calculating the remaining $x_i$'s and $y_i$'s using equation \ref{eq:digitequalspacecondition}. Afterwards one checks whether this choice of values satisfy the conditions given by equation \ref{eq:digitareacondition}. If that is the case, the $x_i$'s and $y_i$'s were chosen appropriately.

@@

\begin{equation} \label{eq:1bit}
\hat{s}_1(t) = \left\{ \begin{array}{lr}
1, & \text{for } s(t) > 0\\
-1, & \text{for } s(t) \leq 0
\end{array} \right. \end{equation}

@@

\begin{equation}  \label{eq:2bit}
\hat{s}_2(t) = \left\{ \begin{array}{rl}
1.51 \sigma, & \text{for } s(t) \geq 0.9816 \sigma\\
0.4528 \sigma, & \text{for } 0 \leq s(t) < 0.9816 \sigma\\
-0.4528 \sigma, & \text{for } -0.9816 \sigma \leq s(t) < 0\\
-1.51 \sigma, & \text{for } -0.9816 \sigma < s(t)\\
\end{array} \right. . \end{equation}
%C: should i round to equal significant figures here? Unfortunately these are the exact values that i used...

@@

\begin{equation}  \label{eq:3bit}
\hat{s}_3(t) = \left\{ \begin{array}{rl}
2.152 \sigma, & \text{for } s(t) \geq 1.748 \sigma\\
1.344 \sigma, & \text{for } 1.05 \sigma \leq s(t) < 1.748 \sigma\\
0.756 \sigma, & \text{for } 0.501 \sigma \leq s(t) < 1.05 \sigma\\
0.245 \sigma, & \text{for } 0 \leq s(t) < 0.501 \sigma\\
-0.245 \sigma, & \text{for } -0.501 \sigma \leq s(t) < 0\\
-0.756 \sigma, & \text{for } -1.05 \sigma \leq s(t) < -0.501 \sigma\\
-1.344 \sigma, & \text{for } -1.748 \sigma \leq s(t) < -1.05 \sigma\\
-2.152 \sigma, & \text{for } -1.748 \sigma < s(t)\\
\end{array} \right. . \end{equation}

The fact that the single bit digitisation scheme makes no references to the standard deviation of the signal is irrelevant. This affects the normalisation of the signal, which we can account for during later analysis as demonstrated in the appendix. What is important is the spacing between digitisation thresholds and relative distance between the output levels chosen.

%Completely different digitisation schemes can be thought of, which would for example reference the most drastic outlier in the data set, or seek to place equal numbers of points into each digitisation level. However, given the assumptions made the schemes derived above minimise the distortion metric defined. Additionally, they are simple enough to be easily implemented computationally. We consider how to design digitisation schemes that preserve the power of the TOD and how to adjust a given digitisation scheme to ensure power conservation in the appendix.

@@

To investigate the performance of the derived digitisation schemes we simulate many scans over CMB template maps at the timestream level. We obtain control maps that use 64bit TOD and maps that have undergone 1, 2 and 3 bit digitisation at the timestream level. We calculate the temperature and polarisation power spectra of each map and determine the noise level.
%C: Do I need to mention that we assume to have an infinitely fast spinning plate?

To create the template maps we use the Hierarchical Equal Area isoLatitude Pixelization of a sphere \citep[HEALPix;][]{healpix}\footnote{Available at \url{http://healpix.sourceforge.net/}} framework. We generate a realisation of I, Q and U maps with $\mathrm{NSIDE} = 4096$ based on the 2015 results of the Planck collaboration. The key cosmological parameters are $\Omega_b \approx 0.049 $, $\Omega_c \approx 0.265$, $\Omega_m \approx 0.316$ and $h \approx 0.67$ \citep{planck2016}.

@@

We simulate observing a $\sim 600 \mathrm{deg^2}$ patch of the sky with a single detector. To do so we perform a number of constant elevation scans (CES), equally spaced in declination (DEC). We repeat the observation strategy $100$ times with a slight offset in right ascension (RA) and DEC each time, such that each pixel is sampled approximately uniformly. The detector is read out at $200 \mathrm{Hz}$. The speed at which we sweep across the survey area is adjusted to produce a desired number of hits per pixel in the output maps.

While performing each CES the pixels being targeted are determined. The corresponding values from the template maps are accessed and added to detector noise realisations. For temperature observations we assume the detector noise level to be $500 \mathrm{\mu K \sqrt{s}}$ and for polarisation observations $\sqrt{2} \hspace{0.7ex} 500 \mathrm{\mu K \sqrt{s}}$, i.e. the corresponding photon noise limit.

Having constructed I,Q and U TOD we apply the digitisation schemes derived above. We compress the timestreams into maps by averaging all hits falling into the same pixel. We produce 24 CMB maps in total: three control maps and nine maps with three each obtained from the different digitisation schemes. We simulate six observations, producing maps with $800$, $8,000$, $80,000$, $1,024,000$, $10,240,000$ and $102,400,000$ hits in each pixel.

%It is often commonplace to fit a polynomial to the TOD, a series of sines and cosines, subtract local averages or do a combination of these techniques \citep[e.g.][, planck?]{chown2018} to reduce $1/f$ noise. We do not employ any of these filtering mechanisms. A thorough application of such noise removal techniques would lead to similar results as the white noise case.

@@

%The calculation outlined above has considerable computational requirements if we want to reach up to $\sim 10^8$ hits per pixel. This calculation was made possible through use of the NERSC facilities.

@@

We use the Spatially Inhomogeneous Correlation Estimator for Temperature and Polarisation \citep[PolSpice;][]{polspice} to compute $\mathrm{TT}, \mathrm{EE}$ and $\mathrm{BB}$ power spectra. To minimise cut sky effects we apodise the observed path using a cosine mask with $\sigma_{\mathrm{APOD}} = \sqrt{600}/2 \hspace{0.7ex} \mathrm{deg}$. %C: Should I also mention the multiplication with a cosine mask that I supply in addition to setting this option on polspice? This will likely be reworked when using Healpix to get the power spectra though.

@@

Normalisation of the obtained power spectra is necessary. The digitisation schemes introduced in section \S\ref{sec:dig} are designed to minimise the distortion of the TOD, not to preserve power. We obtain the appropriate normalisation as described in the appendix.

%Normalising obtained spectra against previous experiments is a common calibration technique in many CMB experiments (e.g. SPT). We again refer to the discussion in the appendix on how to modify digitisation schemes to ensure power preservation.%C: Do you know a good reference for this?

All digitisation schemes are able to recover the input power spectra well. This is shown by figure \ref{fig:psrecover}.

@@

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/psrecovery.pdf}
  \caption[Current ]{
  Input plus detector noise power spectra (black) and power spectra using 3-bit TOD (violet) for observations with $\sim 10^6$ hits per map pixel. Differences occur because we observe a single CMB realisation.
\label{fig:psrecover}
}
\end{figure}

\subsection{White Noise}
\label{subsec:whitenoise}

To investigate whether the noise added through digitisation has any $\ell$ dependence we subtract the template maps from the simulated observations. We use theoretically calculated normalisation constants to minimise calibration error. The power spectra of the difference maps do not show any $\ell$ dependence in the added noise as seen in figure \ref{fig:diffpswn}. % This is in agreement with (planck website) but does not match the results of (planck short paper). This is likely due to the intricities of the planck compression scheme differing to what is considered here. A full comparison is difficult in the absense of a detailed description of the simulations carried out in either paper.

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/diffwn.pdf}
  \caption[Current ]{
  Power spectra of reconstructed minus input maps for 1 Bit (green), 2 Bit (orange), 3 Bit (violet) and the control case (black). We do not observe any $\ell$ dependence. Simulations with $\sim 10^6$ hits per map pixel are shown.
\label{fig:diffpswn}
}
\end{figure}

To quantify the distortion caused by extreme digitisation we infer the map noise level from the relevant power spectra and compare them to the map noise level of control observations. We formulate

@@

In this work we have motivated the investigation of extreme digitisation as a technique in combating potential challenges of CMB experiments. The reduction of TOD by an order of magnitude addresses potential issues in transmission from remote locations. Benefits in mission planning, data analysis and hardware requirements are possible.
